---
title: '<font size="5">Sublime Text 3</font><br>Packages'
output:
    html_document:
        theme: paper
        highlight: textmate
        # number_sections: true
        toc: true
---
<font size="3">

[⇦ Back](../sublime_text.html)

Install Packages
================
To install packages:

- Install **Package Control** (Tools ⭢ Install Package Control...)
- Close and re-open Sublime Text once this is done
- Open Command Palette (Menu bar ⭢ Tools ⭢ Command Palette... or one of the following key bindings:)
    - *macOS:* Shift+Cmd+P
    - *Windows/Linux:* Shift+Ctrl+P
- Search "Install Package" ⭢ Enter ⭢ search for the package you want to install
- Close and re-open Sublime Text after each installation

> **_Note:_** to see where package configuration files are stored, go to Preferences ⭢ Browse Packages...

<!-- *macOS:* /Users/&lt;username&gt;/Library/Application Support/Sublime Text 3/Packages -->

To uninstall packages, go to Command Palette ⭢ search for "Remove Package" ⭢ Enter ⭢ select package to remove

Recommended Packages
====================
*Note:* a lot of these packages have functionality that can be added to key bindings to improve their usability. See [the Key Bindings page](key_bindings.html) for more.

- **Terminus**
    - Allows you to create a terminal (a command-line interface) in the Sublime Text editor
    - To open a terminal in its own panel:
        - Open the Command Palette:
            - *macOS:* Shift+Cmd+P
            - *Windows/Linux:* Shift+Ctrl+P
        - Search for "Terminus: Open Default Shell in Panel" ⭢ Enter
    - To open a terminal as a tab:
        - Open the Command Palette
        - Search for "Terminus: Open Default Shell in Tab (View)" ⭢ Enter
    - To create a key binding that opens a terminal pointing to the same folder that the active file is in:
        - Open Key Bindings:
            - *macOS:* Sublime Text ⭢ Preferences ⭢ Key Bindings
            - *Windows/Linux:* Preferences ⭢ Key Bindings
        - Add the following in the right-hand-side (the User key bindings file):
        ```{json}
        { "keys": ["shift+alt+t"], "command": "terminus_open", "args": {"cwd": "${file_path:${folder}}"} },
        ```
        - The keyboard shortcut Shift+Alt+T should now work
    - *Note:* instead of reading `.bashrc` upon start up, the Terminus terminal reads `.bash_profile`. This means that any customisation you have done to your terminal will not be applied to these Terminus terminals unless you copy this customisation in `.bashrc` into `.bash_profile` too.
        - The most common such customisation is when people re-define their default version of Python from 2 to 3 by adding `alias python=python3` to `.bashrc`
        - If this is you (and it should be - you should not be using Python 2 anymore!) make sure you do this in `.bash_profile` too, eg:
            - Run `sublime ~/.bash_profile` or `subl ~/.bash_profile` in terminal (depending what OS you are on)
            - Appending the following:
            ```{text}
            # the terminal command "python" in sublime text terminals should open python3, not python2
            alias python=python3
            ```
- **LaTeXTools**
    - Allows you to build Latex files in the Sublime Text Build System
    - Latex must be installed on your computer for this to work
    - To use, open a Latex file and invoke the Build command (Ctrl+B, or Cmd+B on macOS)
    - For more info (eg adding a dedicated key binding and/or creating a build command to help troubleshoot if the default build system doesn't work) see [the Latex page](latex.html)
- **LaTeX Word Count**
    + See the word count in a Latex file
    - With a Latex file open, use as follows:
        - *macOS:* Shift+Cmd+C
        - *Windows/Linux:* Shift+Ctrl+C
    - The one drawback is that this does not display the live word count in the status bar. However, the WordCount package does (so install both).
- **WordCount**
    - See the live word count in the status bar
- **Pretty JSON**
    - Make JSON files more readable
    - Key bindings:
        - *macOS:* Ctrl+Cmd+J
        - *Windows/Linux:* Ctrl+Alt+J
- **MarkdownEditing**
    - A general Markdown plugin. See more about using Markdown in Sublime Text [here](markdown.html).
    + *Note:* this package introduces a key binding conflict as it overwrites Ctrl+1/2/3. This isn't an issue if you don't use those bindings, but if you do then create a [user key binding](key_bindings.html) to fix it.
    - To change the colour scheme of Markdown text windows:
        - Have the window with the undesirable colour scheme active
        - Preferences ⭢ Settings - Syntax Specific ⭢ add the following into the Markdown file (right-hand side):
        ```{json}
        {
            "extensions":
            [
                "md"
            ],
            "color_scheme": "Packages/Color Scheme - Default/Monokai.tmTheme",
            "save_on_focus_lost": true,
            "line_numbers": true,
            "translate_tabs_to_spaces": false
        }
        ```
    - If it starts to think that .txt files are Markdown file, open a .txt file (ie have that window active) and go: View ⭢ Syntax ⭢ Open all with current extension as... ⭢ your syntax choice
- **MarkdownPreview**
    - View rendered Markdown in your browser
    - You can create a [user key binding](key_bindings.html) to immediately open the current Markdown file as a rendered HTML page in your web browser
- **LiveReload**
    - Automatically refresh a webpage. This is useful to use in conjunction with MarkdownPreview as you can configure it to update your rendered text every time you save. This allows you to work in Markdown in one window and see your rendered output in another.
    - Use with: Command Palette... ⭢ LiveReload: Enable/disable plug-ins ⭢ Enable - Simple reload
    - Make this the default behaviour so you don't need to turn it on every time you re-open Sublime Text:
        - Open the package settings
            - *macOS:* Menu bar ⭢ Sublime Text ⭢ Preferences ⭢ Package Settings ⭢ LiveReload ⭢ Settings - User
            - *Ubuntu:* Menu bar ⭢ Preferences ⭢ Package Settings ⭢ LiveReload ⭢ Settings - User
        - Add "simple_reload" as in the below:
        ```{json}
        {
            "enabled_plugins": ["SimpleRefresh"]
        }
        ```
- **Markdown Table Formatter**
    - Easily align columns in Markdown tables
    - Default key binding: Shift+Ctrl+Alt+T
- **R-IDE**
    - Render code with (and get syntax highlighting for) R Markdown
    - For more info, see the [R](R.html) and [R Markdown](r_markdown.html) pages
    - You also need to have R Markdown installed on your computer. Run the following in R (either in the RStudio console or in R in the terminal) to do this:
    ```{r, eval = FALSE}
    install.packages('rmarkdown', repos='http://cran.us.r-project.org')
    ```
    - If you want to render Python scripts with R Markdown you also need to have `reticulate` installed. Run the following in R to do this:
    ```{r, eval = FALSE}
    install.packages('reticulate', repos='http://cran.us.r-project.org')
    ```
- **LSP**
    + This package (the Language Server Protocol) needs to be installed for rendering of R Markdown to work
- **Non Text Files**
    - Open PDFs (and other binary files) in external applications, not in Sublime Text
+ **SublimeLinter**
    - Lint your code (ie check it for stylistic and formatting errors)
    + SublimeLinter is a linter *framework* - the linters themselves are not included and must be installed separately, both in Sublime Text on your computer
    + To see the results of the linting every time you save:
        - Open Preferences ⭢ Package Settings ⭢ SublimeLinter ⭢ Settings
        - Add to the user settings (on the right-hand-side): `"show_panel_on_save": "view"`
    + Install the following linters (there's no harm in installing more rather than fewer):
        - Python:
            - Install these linters through pip:
                - `python -m pip install flake8`
                - `python -m pip install pycodestyle`
                - `python -m pip install pydocstyle`
                - `python -m pip install bandit`
                - `python -m pip install pyflakes`
            - Install these corresponding Sublime Text linter packages through Package Control:
                * SublimeLinter-flake8
                * SublimeLinter-pycodestyle
                * SublimeLinter-pydocstyle
                * SublimeLinter-bandit
                * SublimeLinter-pyflakes
            - Make sure you have the Python 3 versions, not Python 2 ones!
        - R:
            + Install the `lintr` package in R: `install.packages("lintr", repos = "http://cran.us.r-project.org")`
            + Install the `SublimeLinter-contrib-lintr` package in Sublime Text (see below)
- **Python Flake8 Lint**
    - Additional linting for Python
- **Python PEP8 Autoformat**
    + Interactively reformat Python code according to PEP8
- **Python Open Module (New)**
    + Opens the Python module file based on the Python import path
- **SublimeLinter-contrib-lintr**
    + This is needed to get the R linter (`lintr`) working
    + This is not always available in Package Control, so you can download it manually from [here](https://github.com/jimhester/SublimeLinter-contrib-lintr). Unzip it and move the folder into your packages folder (find where this is by opening Preferences ⭢ Browse Packages...).
- **INI**
    - Syntax highlighting for .ini files (initialisation/configuration files)
- **FindKeyConflicts**
    + Check to see if you have conflicting key bindings

Other Packages
==============
- **SublimeREPL**
    - Enables code to be run in a separate tab in the editor (as opposed to in a panel at the bottom, which is what the default Sublime Text Build System does)
    - This is neat but it doesn't add to what Sublime Text can already do
    - It also enables you to create a bash terminal as a tab in the editor (via Tools ⭢ SublimeREPL ⭢ Shell), but...
    - ...this opens a Sublime Text version of the terminal, not the actual terminal. This means, eg, that the PATH variable is different
    - Rather use the Terminus package to create terminals in the editor
- **ElasticTabstops**
    - Allows spacing that is created by using Tab to be more intuitive
- **Chain of Command**
    - Execute two commands at once
- **sublimehighlight**
    - Enables the ability to copy text whilst retaining the syntax highlighting
- **MarkdownLivePreview**
    - Use the MarkdownPreview + LiveReload functionality instead
- **AutomaticFullscreen**
    + It got annoying
- **SummitLinter**
    - A different attempt at integrating linting
- **SublimeLinter Inline Errors**
    - This linter does not work well, rather use SublimeLinter
- **Behave Toolkit**
    + Integration with Behave
- **Jira**
    + Integration with Jira
- **Package​Resource​Viewer**
    - View package resources

[⇦ Back](../sublime_text.html)

</font>
