---
title: '<font size="5">Data Handling in R:</font><br>Data Classification Using Date or Time Information'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../google_analytics.html
---

<font size="3">

[⇦ Back](../../R.html)

The Problem
===========
Here we have a data set from a fictitious research study that is *incomplete* - it only specifies when the participants visited the study centre for the first time, and lists these in the 'Visit Number' column as 'Visit 1':

```{r}
df <- data.frame(
    participant_id = c(102, 105, 105, 111, 111, 111),
    visit_date = c(
        "2018-02-20", "2018-04-24", "2019-02-15",
        "2018-01-27", "2013-11-19", "2018-11-28"
    ),
    visit_no = c("Visit 1", "Visit 1", NA, "Visit 1", NA, NA)
)

print(df)
```

What we want to do is fill in the gaps: calculate from the dates which visits were 'Visit 0's and 'Visit 2's. More specifically, we can see that:

- Participant 105 had their 'Visit 1' in April 2018, so their visit in Feb 2019 was their 'Visit 2'
- Participant 111 had their 'Visit 1' in Jan 2018, so their visit in Nov 2013 was their 'Visit 0' and their visit in Nov 2018 was their 'Visit 2'

Let's see how to calculate that automatically:

The Solution
============
For each participant, we want to filter to get only their visit information, get the date of their Visit 1 and then assess, for each other visit, whether it came before or after this date:

```{r}
# Iterate over the participant IDs
for (id in unique(df$participant_id)) {
    # Extract data for this participant ID
    this_id <- subset(df, participant_id == id)
    # Date of visit 1
    visit1 <- subset(this_id, visit_no == "Visit 1")[["visit_date"]]
    # The visit before visit 1 is visit 0
    for (i in rownames(this_id)) {
        date <- df[[i, "visit_date"]]
        if (visit1 > date) {
            df[[i, "visit_no"]] <- "Visit 0"
        } else if (visit1 < date) {
            df[[i, "visit_no"]] <- "Visit 2"
        }
    }
}

print(df)
```

[⇦ Back](../../R.html)

</font>
