---
title: '<font size="5">Data Handling in R:</font><br>Create Unique Identifiers'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../google_analytics.html
---

<font size="3">

[⇦ Back](../../R.html)

<!-- This script needs: -->
<!-- install.packages("R.utils") -->

This page discusses **how to work with data when there is no unique identifier** for each data point.

Let's imagine we're running an experiment with participants in three locations - London, Birmingham and Manchester - who are asked to complete two tests on two different occasions. The raw data, once imported from csv, might look something like this:

```{r}
df <- read.csv("create_unique_identifiers.csv")

library(kableExtra)

kable_input <- kable(df)
kable_styling(
    kable_input, bootstrap_options = c("striped", "scale_down", "condensed"),
    full_width = F, font_size = 11
)
```

A feature of the above data is that there is no unique identifier: no single column contains values that only appear once. There are multiple rows with "London" as the collection site, "A" as the ID, "1" as the timepoint and so on. This makes it difficult to analyse this data; if you want to extract the first participant's data you can't just filter for "Participant A" as there are three Participant As. You need to *create an identifier* manually:

```{r}
# Create a new column with a unique ID for each participant
df["UID"] <- paste0(df$Data_Collection_Site, "-", df$Participant_ID)
print(head(df))

# Create a list of the participants
participants <- unique(df[["UID"]])
print(participants)
```

In this case we've used "UID" for our new column, which stands for "**U**nique **ID**entifier".

You might also have noticed that the format of the results of the two tests is different: the column `Passed_1` contains logical data while `Passed_2` contains binary 1s and 0s. Fortunately, this doesn't matter! R is smart enough to know that `1` and `TRUE` mean the same thing (and that `0` and `FALSE` mean the same thing) when the data needs to be binary.

Now we can count the number of people who passed and failed the first test:

```{r}
# Initialise counters
n_passed1 <- 0
n_failed1 <- 0

# For each participant
for (i in seq_len(length(participants))) {
    # Filter to get the rows that correspond to them
    df_sub <- subset(df, UID == participants[i])

    # Sum the values in column "Passed_1" that correspond to them
    passed <- sum(df_sub[["Passed_1"]], na.rm = TRUE)

    if (passed) {
        # If they passed the first test, increase the counter
        n_passed1 <- n_passed1 + 1
    } else {
        # If they didn't pass the first test, increase the other counter
        n_failed1 <- n_failed1 + 1
    }
}
```

These results can be formatted and printed with the help of the `R.utils` library:

```{r message = F, results = "hold"}
library(R.utils)

# Results
per_passed1 <- n_passed1 / length(participants) * 100
per_failed1 <- n_failed1 / length(participants) * 100
printf("Total ppts:          %s\n", length(participants))
printf("Total passed test 1: %s (%3.2f%s)\n", n_passed1, per_passed1, "%")
printf("Total failed test 1: %s (%3.2f%s)\n", n_failed1, per_failed1, "%")
```

Similarly, we can count the passes and fails for the second test:

```{r message = F, results = "hold"}
# Initialise counters
n_passed2 <- 0
n_failed2 <- 0

# For each participant
for (i in seq_len(length(participants))) {
    # Filter to get the rows that correspond to them
    df_sub <- subset(df, UID == participants[i])

    # Sum the values in column "Passed_2" that correspond to them
    passed <- sum(df_sub[["Passed_2"]], na.rm = TRUE)

    if (passed) {
        # If they passed the first test, increase the counter
        n_passed2 <- n_passed2 + 1
    } else {
        # If they didn't pass the first test, increase the other counter
        n_failed2 <- n_failed2 + 1
    }
}

# Results
per_passed2 <- n_passed2 / length(participants) * 100
per_failed2 <- n_failed2 / length(participants) * 100
printf("Total ppts:          %s\n", length(participants))
printf("Total passed test 2: %s (%3.2f%s)\n", n_passed2, per_passed2, "%")
printf("Total failed test 2: %s (%3.2f%s)\n", n_failed2, per_failed2, "%")
```

And, finally, we can add one more complication by counting those who passed both tests as separate to those who only passed one test:

```{r message = F, results = "hold"}
# Initialise counters
n_passed_1and2 <- 0
n_failed_1or2 <- 0

# For each participant
for (i in seq_len(length(participants))) {
    # Filter to get the rows that correspond to them
    df_sub <- subset(df, UID == participants[i])

    # Sum the values in column "Passed_1" that correspond to them
    passed1 <- sum(df_sub[["Passed_1"]], na.rm = TRUE)

    # Sum the values in column "Passed_2" that correspond to them
    passed2 <- sum(df_sub[["Passed_2"]], na.rm = TRUE)

    # Total
    passed <- passed1 + passed2

    if (passed == 2) {
        # If they passed both tests, increase the counter
        n_passed_1and2 <- n_passed_1and2 + 1
    } else {
        # If they failed one of the tests, increase the counter
        n_failed_1or2 <- n_failed_1or2 + 1
    }
}

# Results
per_passed_1and2 <- n_passed_1and2 / length(participants) * 100
per_failed_1or2 <- n_failed_1or2 / length(participants) * 100
printf("Total ppts:          %s\n", length(participants))
printf("Total passed both:   %s (%3.2f%s)\n", n_passed_1and2, per_passed_1and2, "%")
printf("Total failed either: %s (%3.2f%s)\n", n_failed_1or2, per_failed_1or2, "%")
```

[⇦ Back](../../R.html)

</font>
