---
title: '<font size="5">Data Handling in R:</font><br>Descriptive and Summary Statistics'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../google_analytics.html
---

<font size="3">

[⇦ Back](../../R.html)

```{r, echo = FALSE}
options(width = 120)
```

<!-- This script needs: -->
<!-- install.packages("psych") -->

This page is a more advanced version of the [intro to descriptive statistics](../statistics/descriptive_statistics.html).

Means, Medians, IQRs
====================
Find the basics statistics that describe a set of data points.

Vectors
-------
If your data is stored in a vector, you can simply run functions like `median()`, `IQR()` and so on. Because R is primarily made for statistics, these type of operations are named in a very logical and easy-to-remember way!

```{r, results = "hold"}
data <- c(1, 2, 3, 4, 5, 6, 7, 8, 10, 11)
med <- median(data)
print(med)
avg <- mean(data)
print(avg)
iqr <- IQR(data)
print(iqr)
```

Data Frames
-----------
The columns of a data frame are actually vectors, so the same operations as above are also very straightforward:

```{r, results = "hold"}
df <- data.frame(
    group = c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1),
    data = c(1, 2, 3, 4, 5, 6, 7, 8, 10, 11)
)
med <- median(df$data)
print(med)
avg <- mean(df$data)
print(avg)
iqr <- IQR(df$data)
print(iqr)
```

Do Everything at Once
---------------------
We can calculate all these summary statistics in one go with the help of the `describeBy()` function from the `psych` package. With this function we can just throw in a data frame column and it will output a table with all the information we want, although doing so will generate a warning:

```{r}
library(psych)

# Without specifying a group
# (gives a warning)
db <- psych::describeBy(df$data)
print(db)
```

A single value from this table can be accessed by indexing it:

```{r}
print(db$median)
```

If we want to avoid the warning that the above call to `describeBy()` created, we need to *specify a group.* In our example, we have a data frame where once column is called "group" and it tells us that all the data points are in group "1":

```{r}
# Specifying a group
db <- psych::describeBy(df$data, group = df$group)
print(db)
```

You can see that no warning was generated this time, although in order to access a single value we now need to index the table twice:

```{r}
print(db[[1]]$median)
```

Creating Summary Statistics Tables - one statistics, multiple metrics
=====================================================================

Example Data
------------
For this example we are going to use the "Gapminder" dataset which can be installed with the `gapminder` package:

```{r}
library(gapminder)
```

This dataset contains demographic information about multiple countries across different time points:

```{r, results = "hold"}
# Convert the tibble to a data frame
gapminder <- as.data.frame(gapminder)
# Preview our data
print(head(gapminder, 15))
print(tail(gapminder))
```

This dataset is a bit larger than what we need for this tutorial, so let's trim it down a bit - to 27 rows - just to make it easier to use:

```{r}
df <- subset(
    gapminder,
    (continent == "Oceania" | continent == "Americas") & year == 2007
)
print(df)
```

Establish a 'Discriminator'
---------------------------
A 'discriminator' is like a category. In this example we are going to categorise the 27 countries in our dataset as either having a "Large" or a "Small" population:

```{r}
df$discriminator <- ifelse(df$pop >= 10000000, "Large", "Small")
print(df)
```

Use the `group_by()` function from the `dplyr` package to sort this data into two groups by the discriminator, ie sort the countries into "Large" and "Small":

```{r, message = FALSE}
library(dplyr)

table <- dplyr::group_by(df, discriminator)
```

Calculate Means by Group
------------------------
The `summarise()` function together with the `mean()` function will produce a table that shows the means (and only the means) of each metric (life expectancy, population and GDP per capita) for each group (Large and Small countries):

```{r, message = FALSE}
table_mean <- summarise(
    table,
    lifeExp = mean(lifeExp, na.rm = TRUE),
    pop = mean(pop, na.rm = TRUE),
    gdpPercap = mean(gdpPercap, na.rm = TRUE)
)
# Convert the tibble to a data frame
table_mean <- as.data.frame(table_mean)
print(table_mean)
```

This table is in 'wide' format, but we can convert it into 'long' format using the `gather()` function from the `tidyr` package:

```{r}
library(tidyr)

# Convert from wide to long
table_mean <- tidyr::gather(
    table_mean, key = "variable", "value", lifeExp, pop, gdpPercap
)
print(table_mean)
```

A similar method that leaves us with the same result but *transposed* (so the discriminator forms the column headings):

```{r, message = FALSE}
table_mean <- summarise(
    table,
    lifeExp = mean(lifeExp, na.rm = TRUE),
    pop = mean(pop, na.rm = TRUE),
    gdpPercap = mean(gdpPercap, na.rm = TRUE)
)
table_mean <- as.data.frame(t(table_mean))
colnames(table_mean) <- table_mean[1, ]
table_mean <- table_mean[-1, ]
print(table_mean)
```

Format the Output
-----------------
We can change what the table looks like with a bit of string formatting (using `sprintf()`):

```{r, message = FALSE}
table_mean <- summarise(
    table,
    `Life Exp` = sprintf("%4.1f (%3.1f)", mean(lifeExp), sd(lifeExp)),
    `Population` = sprintf("%4.1f (%3.1f)", mean(pop) / 10^6, sd(pop) / 10^6),
    `GDP per Capita` = sprintf("%4.1f (%3.1f)", mean(gdpPercap), sd(gdpPercap))
)
table_mean <- as.data.frame(t(table_mean))
colnames(table_mean) <- table_mean[1, ]
table_mean <- table_mean[-1, ]
print(table_mean)
```

Creating Summary Statistics Tables - multiple statistics, one metric
====================================================================
This time we are creating a table with multiple statistics (instead of just the mean), but just for the 'population' metric:

```{r, message = FALSE}
table_pop <- summarise(
    table,
    median = median(pop, na.rm = T),
    n = n(),
    non_na_count = sum(!is.na(pop)),
    stdev = sd(pop, na.rm = T)
)
table_pop <- as.data.frame(table_pop)
print(table_pop)
```

A specific value can be extracted from the table through indexing:

```{r}
print(table_pop[1, "median"])
print(table_pop[table_pop$discriminator == "Small", "median"])
```

Here it is transposed and formatted:

```{r, message = FALSE}
table_pop <- summarise(
    table,
    Median = sprintf("%3.1f million", median(pop) / 10^6),
    n = n(),
    `Non-nulls` = sum(!is.na(pop)),
    `Standard deviation` = sprintf("%3.1f million", sd(pop) / 10^6)
)
table_pop <- as.data.frame(t(table_pop))
colnames(table_pop) <- table_pop[1, ]
table_pop <- table_pop[-1, ]
print(table_pop)
```

[⇦ Back](../../R.html)

</font>
