---
title: '<font size="5">Data Handling in R:</font><br>Filter by Number of Times a Values Appears'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../google_analytics.html
---

<font size="3">

[⇦ Back](../../R.html)

If you are running an experiment with multiple timepoints and you want to **select only the paired samples** - ie the ones that have been tested twice - you need to filter by the number of times a values appears. Take the following example data:

```{r}
df <- data.frame(
    id = c(
        "101-0001", "101-0001", "101-0002", "101-0002", "101-0003", "101-0004",
        "101-0005"
    ),
    timepoint = c("A", "B", "A", "B", "A", "A", "A")
)
print(df)
```

We can see that participants "101-0001" and "101-0002" have been tested at two timepoints each. To select only them, you can group by `id` then file for those that appear twice using dplyr:

```{r}
library(dplyr, warn.conflicts = FALSE)

df %>%
    group_by(id) %>%
    filter(n() == 2) -> subset

print(subset)
```

...or, using only Base R, you can split the data frame `by` factors and count the number of rows for each factor using `nrow`. Then, search for factors that appear twice and filter by that search result:

```{r}
res <- by(df, df$id, nrow)
res <- res == 2
subset <- df[df$id %in% names(which(res)), ]

print(subset)
```

[⇦ Back](../../R.html)

</font>
