---
title: '<font size="5">Data Handling in R:</font><br>Pivot Tables'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../google_analytics.html
---

<font size="3">

[⇦ Back](../../R.html)

```{r, echo = FALSE}
options(width = 120)
```

<!-- This script needs: -->
<!-- install.packages("tidyr") -->
<!-- install.packages("gapminder") -->
<!-- install.packages("reshape") -->
<!-- install.packages("reshape2") -->

A "pivot table" is a way of summarising the data in a table. For example, if we have the following results of three tests that were taken by two students:

```{r, echo = FALSE}
results <- data.frame(
    Student = c("A", "B", "A", "B", "A", "B"),
    Test = c("Test 1", "Test 1", "Test 2", "Test 2", "Test 3", "Test 3"),
    Score = c(66, 76, 80, 77, 59, 61)
)
library(kableExtra)
kable_input <- kable(results)
kable_styling(
    kable_input, bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
)
```

Then a pivot table would look as follows:

```{r, echo = FALSE}
library(tidyr)
wide <- spread(results, Test, Score)
kable_input <- kable(wide)
kable_styling(
    kable_input, bootstrap_options = c("striped", "hover", "condensed"),
    full_width = F
)
```

As you can see, the first table was **long** while the second was **wide**. That's another way of talking about this type of operation: going from long to wide or, if you want to go the opposite way, from wide to long.

From Long to Wide
=================
For this example we will use the ["Gapminder" dataset](https://cran.r-project.org/web/packages/gapminder/readme/README.html), which can be installed as follows:

```{r, eval = FALSE}
install.packages("gapminder", repos = "http://cran.us.r-project.org")
```

This dataset contains the life expectancy, population and GDP per capita of 142 countries as recorded in various years:

```{r, results = "hold"}
# Import the gapminder as a tibble
library(gapminder)
# Convert tibble to data frame
gapminder <- as.data.frame(gapminder)
# Display
head(gapminder, n = 20)
cat("\n\t\t\t\t\t\t\t...\n\n")
tail(gapminder)
```

As you can see, this is a **long** dataset. To pivot it and make it **wide**, you have two options:

Using spread() from tidyr
-------------------------
The `spread()` function allows us to choose which variable to use as the column headings and which to use as the data in the table:

```{r}
library(tidyr)
# Select the columns we are interested in
long <- gapminder[c("country", "continent", "year", "pop")]
# Pivot
wide <- spread(long, year, pop)
```

Show the population in each country in each year sampled:

```{r}
head(wide[, 1:12])
```

Using cast() from reshape
-------------------------
The `cast()` function allows us to do the same thing without needing to first trim down to just the columns we want:

```{r, message = F}
library(reshape)
wide <- cast(gapminder, country ~ year, value = "pop")
head(wide[, 1:12])
```

Of course, we do want to keep the "continent" column there:

```{r, message = F}
wide <- cast(gapminder, country + continent ~ year, value = "pop")
head(wide[, 1:12])
```

Using dcast() from reshape2
---------------------------
The reshape2 package has the `acast()` and the `dcast()` functions: `acast()` returns its output as a vector, matrix or array while `dcast()` returns it as a data frame. They are otherwise identical to `cast()` from the reshape package, although they use `value.var` instead of `value` as the keyword for the value parameter:

```{r, message = F}
library(reshape2)
wide <- dcast(gapminder, country + continent ~ year, value.var = "pop")
head(wide[, 1:12])
```

Something that's useful to remember is that the default aggregation function for `dcast()` is `length`. In other words, it will return a count of the number of elements relevant to the row and column headings by default. In this example, it gives the number of countries for which we have data at each time point and from each continent:

```{r, message = F}
wide <- dcast(gapminder, continent ~ year)
print(wide)
```

From Wide to Long
=================
Unsurprisingly, this is the opposite operation: going from a **wide** table to a **long** one.

Using gather() from tidyr
-------------------------
```{r}
# Either explicitly define which columns to use
gathercols <- c(
    "1952", "1957", "1962", "1967", "1972", "1977",
    "1982", "1987", "1992", "1997", "2002", "2007"
)
# Or programmatically extract them from the column names
gathercols <- colnames(wide)[3:length(colnames(wide))]

# Convert from wide format to long format
long <- gather(wide, "year", "pop", all_of(gathercols), factor_key = TRUE)
head(long)
```

From Long to Wide via Summarising
=================================
Lastly, we can generate summary statistics for the data in the table at the same time as we are pivoting it. For this example, let's use the `ChickWeight` pre-loaded dataset which details the weights of chicks on different diets at different ages:

```{r}
head(ChickWeight)
```

Here's how we can pivot the table and end up with the *median* weight of each chick over the time points:

```{r}
chick_weights <- aggregate(
    weight ~ Chick + Diet, data = ChickWeight, FUN = median
)
chick_weights$Chick <- as.numeric(as.character(chick_weights$Chick))
chick_weights <- chick_weights[order(chick_weights$Chick), ]
head(chick_weights)
```

[⇦ Back](../../R.html)

</font>
