---
title: '<font size="5">Introduction to R:</font><br>Booleans and If Statements'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../google_analytics.html
---

<font size="3">

[⇦ Back](../../R.html)

# Booleans

A Boolean is something that is either true or false. In R, the values TRUE and FALSE must be typed with capital letters:
```{r}
y <- TRUE
n <- FALSE
```

## Equality

You can also create a Boolean by testing equality. To do this, use two equal signs: `==`
```{r}
# Is 1 equal to 1?
1 == 1
```

Remember that a SINGLE equals sign is used to assign a value to a variable while a DOUBLE equals sign is used to test for equality:
```{r}
# Assign a value of 1 to a
a <- 1
# Assign a value of 1 to b
b <- 1
# Does a equal b?
a == b
```

This means that you can test equality using `==`, generate a Boolean, then assign this Boolean to a variable using `=`:
```{r}
# The statement "(1 == 1)" evaluates to "TRUE" and then we assign that result
# to the variable "bool"
bool <- (1 == 1)
# Is 1 equal to 1?
print(bool)
```

In addition to equality (`==`) you can test if two things are NOT equal using `!=`. In R, the exclamation mark means 'not':
```{r}
# Is 1 not equal to 1?
bool <- (1 != 1)
print(bool)
# Is 1 not equal to 2?
bool <- (1 != 2)
print(bool)
```

## Inequality
You can use `>` and `<` to test if something is greater than or less than something else. Use `>=` to test if something is greater than *or equal to* something else and `<=` to test 'less than or equal to':
```{r}
# Is 1 less than 2?
print(1 < 2)
# Is attendance below or at capacity?
attendance <- 95
capacity <- 100
print(attendance <= capacity)
```

## 'And' and 'Or'

You can combine conditionals using 'and' and 'or' statements.

An 'and' statement is created by using an ampersand (`&`) between two conditionals. In order for an 'and' statement to evaluate to `TRUE` both conditionals must be true:
```{r}
# Is attendance below or at capacity AND
# has everyone bought a ticket?
attendance <- 95
capacity <- 100
tickets_sold <- 93
print((attendance <= capacity) & (tickets_sold == attendance))
```

Notice the use of brackets in the last line. Each conditional is enclosed in round brackets to aid readability.

An 'or' statement is created by using a pipe (`|`) between two conditionals. In order for an 'or' statement to evaluate to `TRUE` either conditional can be true:
```{r}
# Is attendance over capacity OR
# has someone not bought a ticket?
attendance <- 105
capacity <- 100
tickets_sold <- 105
print((attendance > capacity) | (tickets_sold != attendance))
```

Notice that the same question is being asked as in the 'and' statement example, but this time in a different way.

## Strings

Strings are handled similarly to numbers:
```{r}
# Do these two people have the same name?
person1 <- "WINSTON CHURCHILL"
person2 <- "Winston Churchill"
print(tolower(person1) == tolower(person2))
```
Notice the function `tolower()` that was used above to convert both strings to lowercase before making the comparison. Strings are case sensitive, so if we hadn't used that function the result would have been `FALSE`.

You can also test if a string contains a certain sub-string or character using `grepl()`:
```{r}
st <- "Hello World"
print(grepl("Hello", st))
```

'grepl' stands for 'global regular expression print - logical', in case you were wondering.

# If/Else Statements

## If Statements

An 'if' statement is a chunk of code that will execute if a condition is true (and which will not execute if the condition is false). In order to write an if statement you need to:

- Put a Boolean condition inside round brackets
- Put the chunk of code to be executed if the condition is true inside curly brackets:

```{r}
if (1 == 1) {
    print("This will print to console because 1 is equal to 1")
}
```

## Else Statements

An 'else' statement is a chunk of code that will execute if the 'if' statement does not execute:
```{r}
height <- 170
if (height >= 180) {
    print("This person is tall")
} else {
    print("This person is not tall")
}
```

## Else If Statements
An 'else if' statement is like a second 'if' statement. It is a chunk of code that will execute if the first 'if' statement does not evaluate to true but the second one does:
```{r}
height <- 170
if (height >= 180) {
    print("This person is tall")
} else if (height < 180 & height >= 160) {
    print("This person is medium height")
} else {
    print("This person is short")
}
```

## Strings

Besides the `grepl()` function mentioned above, another way to check if a string contains a sub-string is to use the `gregexpr()` (global regular expression) function. This returns the index of a sub-string or `-1` if it cannot find it:
```{r}
st <- "Hello World"
# Search for the letter "x" in the string "Hello World"
idx_x <- unlist(gregexpr(pattern = "x", st))
# If the letter "x" is found, print a message that says that it has been found.
# If the letter "x" is not found, print the opposite message.
if (idx_x == -1) {
    print("The string does NOT contain the letter x")
} else {
    print("The string DOES contain the letter x")
}
```
As expected, the letter "x" was not found in "Hello World".

## Using If Statements to Check Your Code

Imagine you are running some code using a dataset that someone gave you and which contains some errors. For example, maybe they entered someone's birth year as 1890 instead of 1980 and so you calculate their age as being 130 instead of 40. You can use if statements together with the `stop()` function to sanitise your data and check that these types of errors don't occur:
```{r echo = FALSE}
current_year <- 2020
birth_year <- 1980
```
```{r results = "hide"}
age <- current_year - birth_year
if (age <= 110) {
    print("This person is not older than 110 years")
} else {
    stop("This person is older than 110 years. Are you sure this is correct?")
}
```

The above code will cause the script to stop and produce the message:  
```This person is older than 110 years. Are you sure their age is correct?```  
which is a lot better than having it run to completion and generating incorrect results.

[⇦ Back](../../R.html)

</font>
