---
title: '<font size="5">Introduction to R:</font><br>Dictionaries'
output:
    html_document:
        theme: paper
        highlight: textmate
---
<font size="3">

[⇦ Back](../../R.html)

Python Dictionaries in R?
=========================
In Python, users have access to a very useful data structure called 'dictionaries'. They literally let a script __translate__ an object into another object. Take a look at the following Python code:

```{python}
# Python code
english_to_german = {
    "potato": "Kartoffel",
    "ambulance": "Krankenwagen",
    "rabbit": "Hase"
}
translation = english_to_german["potato"]
st = f'The German word for "potato" is "{translation}"'
print(st)
```

Dictionaries are so useful we want to be able to use them in R!

Using a Vector like a Dictionary
================================
Start by initialising a list (ie a vector in 'list mode') and entering the English words:

```{r}
# From here onwards all the code is R
english_to_german <- vector(mode = "list", length = 3)
names(english_to_german) <- c("potato", "ambulance", "rabbit")
```

Now enter the German words, associating each one with the corresponding English word:

```{r}
english_to_german["potato"] <- "Kartoffel"
english_to_german["ambulance"] <- "Krankenwagen"
```

Instead of using the English _word_, we can associate the German word with the relevant _index_ of the list:

```{r}
english_to_german[[3]] <- "Hase"
```

Let's take a look at the result:

```{r}
print(english_to_german)
cat(
    "The German word for \"potato\" is \"", english_to_german$"potato", "\"",
    sep = ""
)
```

Using a Data Frame like a Dictionary
====================================
A data frame is a lot more powerful than a list; for example, we could create our dictionary in a spreadsheet then import that as a data frame. We're not going to do that in this example though, we're going to create a data frame manually instead:

```{r}
df <- data.frame(
    English = c("potato", "ambulance", "rabbit"),
    German = c("Kartoffel", "Krankenwagen", "Hase")
)
english_to_german <- vector(mode = "list")
```

Here's how to convert that data frame into a dictionary once it's in R:

```{r}
for (i in seq_along(df)) {
    key <- df[i, "English"]
    value <- df[i, "German"]
    english_to_german[as.character(key)] <- as.character(value)
}
print(english_to_german)
cat(
    "The German word for \"potato\" is \"", english_to_german$"potato", "\"",
    sep = ""
)
```

[⇦ Back](../../R.html)
</font>
