---
title: '<font size="5">Introduction to R:</font><br>Named Vectors and Lists (aka Dictionaries)'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../google_analytics.html
---

<font size="3">

[⇦ Back](../../R.html)

Python Dictionaries in R?
=========================
In Python, users have access to a very useful data structure called 'dictionaries'. They literally let a script __translate__ an object into another object. Take a look at the following Python code which creates a very simple English-to-German language dictionary:

```{python}
# Python code
english_to_german = {
    "potato": "Kartoffel",
    "ambulance": "Krankenwagen",
    "rabbit": "Hase"
}
translation = english_to_german["potato"]
st = f'The German word for "potato" is "{translation}"'
print(st)
```

Dictionaries are so useful - for many aspects of coding, not just for language translation - that we want to be able to use them in R too!

Using a Vector like a Dictionary
================================
While R does not have dictionary objects in the same way as Python it does have the ability to *add names to the elements of a vector.* Remember that a vector is a collection of objects in a particular order were each object can be indexed by referring to its position:

```{r}
# From here onwards all the code is R

# This is a vector:
german <- c("Kartoffel", "Krankenwagen", "Hase")
# Vectors are indexed using the numerical position of the element you want:
german[2]
```

To add names to each of the elements, use the `names()` function:

```{r}
english <- c("potato", "ambulance", "rabbit")
names(german) <- english
```

You now have the ability to index the vector using *either* numerical positions *or* elements' names (although now you have to use TWO square brackets to do so because the vector has two dimensions of information):

```{r}
# Index using numerical position
german[[2]]
# Index using name
german[["potato"]]
```

Let's take a look at the result:

```{r}
print(german)
cat(
    "The German word for \"potato\" is \"", german[["potato"]], "\"", sep = ""
)
```

Example: A Contacts List
------------------------
It's not just literal language dictionaries that the dictionary 'data type' is useful for, here's an example of a list of contacts that stores information about a group of people:

```{r}
# Initialise the data for all your contacts
people <- c("Alice", "Bob", "Carol")
phone_numbers <- c("07410123456", "07420123456", "07430123456")
ages <- c(21, 22, 23)
birthday <- c("25 July", "18 October", "1 April")
# Associate the people with the information to create "dictionaries"
names(phone_numbers) <- people
names(ages) <- people
names(birthday) <- people
```

Now you can look up information about each person just by using their name:

```{r}
phone_numbers[["Alice"]]
```

Example: A Contacts List in Reverse
-----------------------------------
Instead of storing information in variables named for the information type, you can store it in variables names for the *person:*

```{r}
# Initialise the data for all your contacts
information <- c("Phone Number", "Age", "Birthday")
Alice <- c("07410123456", 21, "25 July")
Bob <- c("07420123456", 22, "18 October")
Carol <- c("07430123456", 23, "1 April")
# Associate the information with the people to create "dictionaries"
names(Alice) <- information
names(Bob) <- information
names(Carol) <- information
```

Now you can look up information about each person just by using their vector:

```{r}
Alice[["Phone Number"]]
```

This is the reverse of how we did it in section 2.1. One type of implementation might be more sensible than the other for your particular task.

Vectors vs Lists
================
Note that there is a different data type in R called a **list**. These are *special types of vectors* as reflected by the fact that they can be created using the `vector()` function where the "mode" is specified as being a list:

```{r}
# Initialise a list
english_to_german <- vector(mode = "list", length = 3)
```

Dictionaries can be made out of lists in a slightly different way to vectors. Start by entering the names (the English words):

```{r}
names(english_to_german) <- c("potato", "ambulance", "rabbit")
```

Then enter the values (the German words), associating each one with the corresponding name:

```{r}
english_to_german["potato"] <- "Kartoffel"
english_to_german["ambulance"] <- "Krankenwagen"
```

Instead of using the English *words*, we could alternatively associate the German words with the relevant *indexes* of the list:

```{r}
english_to_german[[3]] <- "Hase"
```

This 'named list' is very similar to the 'named vector' we created in the first example, except we can now also index it with dollar sign notation:

```{r}
# Index using numerical position
english_to_german[[2]]
# Index using name
english_to_german[["potato"]]
# Index using dollar sign notation
english_to_german$"rabbit"
```

There is another important difference between using vectors and lists, which is demonstrated in the following example:

Example: A Dictionary of Data Frames
------------------------------------
It's not just string and numerical data that can be put into a dictionary, anything can be. Here's a example with entire data frames as the elements, specifically, the results of three tests given to a class of students:

```{r}
test1 <- data.frame(
    Student = c("Alice", "Bob", "Carol"),
    Result = c(60, 70, 80)
)
test2 <- data.frame(
    Student = c("Bob", "Carol", "Doug"),
    Result = c(61, 71, 81)
)
test3 <- data.frame(
    Student = c("Alice", "Carol", "Doug"),
    Result = c(62, 72, 82)
)
```

If we try to create a dictionary out of these data frames using a vector we get an unexpected result:

```{r}
# Attempt to create a dictionary
class_tests <- c(test1, test2, test3)
names(class_tests) <- c("Test 1", "Test 2", "Test 3")
class_tests
```

What has happened is that the data frames were appended to one another when the vector was created, meaning that there are now six elements (each of which is a series) as opposed to three (each being a data frame) like we want. We instead need to use a list:

```{r}
# Create a dictionary
class_tests <- list(test1, test2, test3)
names(class_tests) <- c("Test 1", "Test 2", "Test 3")
class_tests
```

Now we can look up the results of any test in particular using its name:

```{r}
class_tests[["Test 3"]]
```

We could get even more crazy and have data frames within dictionaries within vectors within lists within data frames and so on!

Convert Data Frames into Dictionaries
=====================================
A data frame is a lot more powerful than a vector. For example, we could create a dictionary in a spreadsheet then import that into R as a data frame. We're not going to do that in this example though, we're just going to create a data frame manually:

```{r}
# Create a data frame
df <- data.frame(
    English = c("potato", "ambulance", "rabbit"),
    German = c("Kartoffel", "Krankenwagen", "Hase")
)
# Initialise a vector that will become a dictionary
english_to_german <- vector()
```

Here's how to convert that data frame into a dictionary once it's in R:

```{r}
for (i in seq_len(nrow(df))) {
    key <- df[i, "English"]
    value <- df[i, "German"]
    english_to_german[as.character(key)] <- as.character(value)
}
print(english_to_german)
cat(
    "The German word for \"potato\" is \"", english_to_german[["potato"]],
    "\"", sep = ""
)
```

[⇦ Back](../../R.html)
</font>
