---
title: '<font size="5">Introduction to R:</font><br>Packages and Libraries'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        # toc: true
        includes:
            before_body: ../../google_analytics.html
---

<font size="3">

[⇦ Back](../../R.html)

# Synopsis
- A 'package' is the same as a 'library'
- Increase the number of functions you can call by installing __*packages*__
- Once installed, you still need to __*import*__ the library into your script

# What are Packages and Libraries, and Why Do I Need Them?
When you download and install R or RStudio you get access to what is known as __*base*__ R. It's very powerful and can do a lot of things but, as the name suggests, it's very basic. In order to add functionality and be able to use other functions you need to __*install packages*__ and __*import libraries*__.

> **What's the difference between a library and a package?**  
> Nothing. They are two names for the same thing. For some reason you *install* a package and *import* a library but I don't know why.

# Install a Package
Packages can be installed by running the `install.packages()` function. Specify the name of the package you want to install as a string and the repository you want to install it from by using the `repos` __*keyword argument*__:

```{r message=FALSE, warning=FALSE, results='hide', eval=FALSE}
install.packages("BlandAltmanLeh", repos = "http://cran.us.r-project.org")
install.packages("chron", repos = "http://cran.us.r-project.org")
install.packages("DiagrammeR", repos = "http://cran.us.r-project.org")
install.packages("dplyr", repos = "http://cran.us.r-project.org")
install.packages("ggplot2", repos = "http://cran.us.r-project.org")
install.packages("ggpubr", repos = "http://cran.us.r-project.org")
install.packages("kableExtra", repos = "http://cran.us.r-project.org")
install.packages("knitr", repos = "http://cran.us.r-project.org")
install.packages("lintr", repos = "http://cran.us.r-project.org")
install.packages("logging", repos = "http://cran.us.r-project.org")
install.packages("maps", repos = "http://cran.us.r-project.org")
install.packages("matrixStats", repos = "http://cran.us.r-project.org")
install.packages("Rmisc", repos = "http://cran.us.r-project.org")
install.packages("R.utils", repos = "http://cran.us.r-project.org")
install.packages("readr", repos = "http://cran.us.r-project.org")
install.packages("reticulate", repos = "http://cran.us.r-project.org")
install.packages("rmarkdown", repos = "http://cran.us.r-project.org")
install.packages("stringi", repos = "http://cran.us.r-project.org")
install.packages("stringr", repos = "http://cran.us.r-project.org")
install.packages("titanic", repos = "http://cran.us.r-project.org")
install.packages("utf8", repos = "http://cran.us.r-project.org")
```

The "cran" part of the repos URL refers to "CRAN", the Comprehensive R Archive Network where most R packages are stored online.

> You only need to install a package once on your computer. After it's finished installing you can delete the `install.packages()` functions from your R script.

You can find out where on your computer the packages are installed by running `.libPaths()` and see what packages you've installed with `print(as.data.frame(installed.packages()))`.

Updating R
----------
Sometimes you can only download a package if it's compatible with the version of R you have installed. In general, you always want to work with the latest version of R and latest versions of packages. If you're on a macOS or Linux machine (eg Ubuntu) you can check what version of R you have installed from the terminal with:

```{bash, eval = FALSE}
R --version
```

To update R manually, download and install the latest version from [the R website](https://www.r-project.org/). When you re-start RStudio it should automatically use this latest version. Note that if you are moving to a new major version (eg from version 3.5 to version 4.0) then you will have to re-install all your packages. Trying to simply update them will not work, you actually need to re-install them. The normal methods of package installation should work fine so try them first, but if you are having issues the easiest thing is to delete all of your package files completely and then re-install from scratch. To find where your package files are stored on your computer run `.libPaths()` in R and open the folder that is returned. Delete all the folders in there.

> On Ubuntu the folder is usually `/usr/local/lib/R/site-library`  
> On macOS the folder is usually `/Library/Frameworks/R.framework/Versions/<version>/Resources/library`

If you are using Ubuntu, you have the option of updating R from the terminal (there is a package called `updateR` that allows you to update R from the terminal on macOS but it doesn't always work). Instructions on how to do this are below. Note that you will still need to re-install your packages if you are moving to a new major version number.

### Updating R from the Terminal on Ubuntu
If you are using Ubuntu and aren't using RStudio, update R by first opening `/etc/apt/sources.list` and adding whichever of the following lines corresponds to your Ubuntu version:

```{bash eval = FALSE}
# 18.04
deb https://cloud.r-project.org/bin/linux/ubuntu bionic-cran40/
# 16.04
deb https://cloud.r-project.org/bin/linux/ubuntu xenial-cran40/
# 14.04
deb https://cloud.r-project.org/bin/linux/ubuntu trusty-cran40/
```

The above are relevant for R version 4.0. Change the `...-cran40/` part to whatever version numbers you are interested in. Then run the following:

```{bash eval = FALSE}
sudo apt-get update
sudo apt-get install r-base r-base-dev
# To update any R libraries installed via APT:
sudo apt-get upgrade
```

# Import a Library
When you want to use functionality that is part of a particular package, you need to import the package as a library in the script.

**Note:** it's best practice is to import all the libraries you need in a script right in the beginning, in alphabetical order:

```{r message = FALSE}
library(dplyr)
```

You can now use the functionality that is included with the libraries you imported:

```{r}
# Import the 'dplyr' library
library(dplyr)
# As an example, you can now use the 'pipe' operator which puts a value into
# the front of the next pair of brackets:
x <- 49
square_root <- x %>% sqrt()
# This will be the square root of 49:
square_root
```

# Call a Function From a Library
You can call a function that comes from a particular library in one of two ways. Here's an example with the `str_length()` function, which comes from the 'stringr' library and finds the number of characters in a string. The first way is to just call the function as you would any other:

```{r}
# Import the library
library(stringr)
# Find the length of a string of text
length <- str_length("Hello, World")
# Output to the console
print(length)
```

While this is the simplest way of doing it, it can sometimes get confusing as to which function came from which library. This is especially true if you have imported a lot of them and are using all sorts of different functions all over the place. Consider using the alternative method, which involves using two colons to explicitly show which library the function came from:

```{r}
# Find the length of a string of text
length <- stringr::str_length("Hello, World")
# Output to the console
print(length)
```

# Alternative Methods of Importing Libraries
If your script is going to be run on someone else's computer and you don't know if they will have a particular package installed, import it into your script with the `require()` function. If the package is installed, it will import it. If it isn't, it will first download and install it, then import it.

```{r message = FALSE}
require(ggplot2)
require(logging)
```

Sometimes, the messages that get printed to the console when a package is installed or imported get annoying. Suppress them with `suppressPackageStartupMessages()`:

```{r}
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(library(logging))
```

# Summary
- A 'package' is the same as a 'library'
- Increase the number of functions you can call by installing __*packages*__
- Once installed, you still need to __*import*__ the library into your script

[⇦ Back](../../R.html)

</font>
