---
title: '<font size="5">Introduction to R:</font><br>Strings and Characters'
output:
    html_document:
        theme: paper
        highlight: textmate
---
<font size="3">
[⇦ Back](../../R.html)

```{r message=FALSE}
# Import packages
library(R.utils)
library(stringr)
```

## Concatenate Strings
There are a number of different ways to do this, depending on the exact functionality you want.

`paste()` uses a space for separation by default:
```{r}
st = paste('Hello,', 'World')
print(st)
```

This default behaviour can be changed:
```{r}
st = paste('Hello', 'World', sep=', ')
print(st)

st = paste('Hello,', 'World', collapse=NULL)
print(st)
```

`cat()` also uses a single space by default:
```{r}
cat('Hello,', 'World')

cat('Hello,', 'World', sep=" ")
```

`c()` is similar to `cat()` in that it also combines/concatenates objects, but `c()` turns them into *vectors*:
```{r}
st = c('Hello', 'World')
print(st)
```

`paste0()` uses no space for separation by default:
```{r}
st = paste0('Hello,', 'World')
print(st)
```

## Split Strings

```{r}
st = 'Split the words in a sentence.'
st = strsplit(st, ' ')
print(st)
```

## Search for a Character or Substring

```{r}
# Find the indices where certain characters are
st = 'Hello, World'
idx_o = unlist(gregexpr(pattern='o', st))
print(idx_o)
# Remove the text before a certain character
sub_st = substr(st, start=idx_o[1], stop=nchar(st))
print(sub_st)
# Remove the text after a certain character
sub_st = substr(st, start=1, stop=idx_o[1])
print(sub_st)
```

## Lookup the Characters at Particuar Indices

```{r}
st = 'Hello World'
st = substr(st, start=3, stop=4)
print(st)
```

## Replace a Substring

```{r}
st = 'Hello, World'
st = gsub('ello', 'i', st)
print(st)
```

## Insert and Overwrite Characters

```{r}
st = 'Hello World'
str_sub(st, 6, 6) = '_'
str_sub(st, 8, 7) = 'O'
print(st)
```

## Formatted Output

```{r}
x = 1
printf('x = %.2f\n', x)

# Print to console immediately
st = 'Hello world'
printf('Output text is: %s\n', st)
# Write the value to a variable
st = 'Hello world'
st = sprintf('Output text is: %s', st)
print(st[1])
# Using formatted output with a data frame
df = data.frame(
    name = c('Alfa', 'Bravo', 'Charlie')
)
printf('Output text is: %s\n', df$name[1])
```

## Unicode and Emojii

```{r}
st = '\U00B5'  # mu
print(st)
st1 = '\U2705'  # white heavy check mark
st2 = '\U274c'  # cross mark
st3 = '\U0001f609'  # winking face
utf8::utf8_print(c(st1, st2, st3))
```

[⇦ Back](../../R.html)

</font>
