---
title: '<font size="5">Introduction to R:</font><br>Strings and Characters'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../google_analytics.html
---

<font size="3">

[⇦ Back](../../R.html)

Concatenate Strings
===================
There are a number of different ways to combine strings depending on the exact functionality you want:

- Use `paste()` or `paste0()` if you want a **character** object to be returned
- Use `c()` if you want a **vector** to be returned
- Use `cat()` if you want nothing to be returned (ie you just want the string to **print to screen**)

Return a Character Object
-------------------------
`paste()` combines strings by adding a space in between them. The result (which is a character object) can then be assigned to a variable:

```{r}
st <- paste("Hello", "World")
print(st)
```

The default behaviour of adding a space in between each string can be changed by using the `sep` keyword argument. For example, you can have a comma and a space between each string by doing the following:

```{r}
st <- paste("Hello", "World", sep = ", ")
print(st)
```

If you want to have **no space** between your strings you can either use `paste()` with no separation or you can use `paste0()` which does this by default:

```{r}
st <- paste("Hello", "World", sep = "")
print(st)
```

```{r}
st <- paste0("Hello", "World")
print(st)
```

Return a Vector
---------------
The `c()` command stands for "**C**oncatenate" and is similar to `paste()` in that it also combines/concatenates objects and allows you to assign them to variables, but `c()` turns them into **vectors** (a list of objects) instead of characters (a single object):

```{r}
st <- c("Hello", "World")
print(st)
```

Print with no Return
--------------------
The `cat()` command is also short for "con**CAT**enate", also performs the same function as `paste()`, also uses a single space by default and can also use the `sep` keyword argument to change this default:

```{r}
cat("Hello", "World")
```

The difference between `paste()` and `cat()` is that `cat()` does not return a value, it only prints to the console. This means that it cannot be saved to a variable like this:

```{r results = "hide"}
a <- cat("Hello", "World")
```

The code below prints `NULL` because nothing was returned by the `cat()` function and so nothing was saved to the `a` variable:

```{r}
print(a)
```

Split Strings
=============
Use `strsplit()` to do the opposite of concatenation:

```{r}
st <- "Split the words in a sentence."
st <- strsplit(st, " ")
print(st)
```

Search, Find, Lookup
====================
You can:

- **Search** a string to see if it contains a certain letter or sub-string
- **Find** where that letter or sub-string is located within the string
- **Lookup** what letter or sub-string is at a certain location within a string

Search for a Character or Sub-string
------------------------------------
Does a string contain the letter(s) you are looking for? Search the string and return a Boolean (true or false) to find out. This is done with the `grepl()` command:

```{r}
# Does the letter "o" appear in "Hello World"?
grepl("o", "Hello World", fixed = TRUE)
```

In case you were wondering, this function's name is short for "**g**lobally search for a **r**egular **e**xpression and **p**rint matching lines - **l**ogical". You can now see why we shorten it!

You may have noticed that we set the keyword argument `fixed` to `TRUE` - this causes the `grepl()` function to search for *exactly* the string that was provided. If this is omitted the default behaviour is to treat the string being searched for as a *regular expression,* which may create unexpected results.

### Search for a Character or Sub-string at the Start or End of a String
Does a string start with or end with a certain thing?

```{r}
startsWith("alphabet", "a")
endsWith("Filename.csv", ".csv")
```

Find a Character or Sub-string
------------------------------
Now that we know our string contains the letter we are looking for, get the **indices** (positions) where that certain character is by using the `gregexpr()` command:

```{r}
st <- "Hello World"
idx_o <- unlist(gregexpr(pattern = "o", st))
print(idx_o)
```

In this example, the letter "o" can be found at both position 5 and 8 in the string "Hello World".

Lookup the Characters at Particular Indices
-------------------------------------------
You can see what characters are at a certain location within a string by using the `substr()` command. This takes the `start` and `stop` keyword arguments, which are pretty self-explanatory (they tell the command where to start and stop looking):

```{r}
st <- "Hello World"
sub <- substr(st, start = 6, stop = 8)
print(sub)
```

You can lookup all the characters from a given start point to the end of a string by using the total number of characters in a string (`nchar()`) as the stop point:

```{r}
sub <- substr(st, start = 6, stop = nchar(st))
print(sub)
```

Similarly, you can lookup all the characters from the start of a string until a given stop point by using "1" as the start point:

```{r}
sub <- substr(st, start = 1, stop = 8)
print(sub)
```

Delete, Overwrite, Replace
==========================
You can:

- **Delete** all the characters before or after a given point in a string
- **Overwrite** the characters at a given location within a string
- **Replace** all occurrences of a certain letter or sub-string within a string with a given replacement

Delete Characters
-----------------
This is again done using the `substr()` function but this time we are looking up text before and after a *character*, not an *index*.

Remove the text *before* a certain character:

```{r}
st <- "Hello World"
idx_o <- unlist(gregexpr(pattern = "o", st))
sub <- substr(st, start = idx_o[1], stop = nchar(st))
print(sub)
```

Remove the text *after* a certain character:

```{r}
sub <- substr(st, start = 1, stop = idx_o[1])
print(sub)
```

As a more practical example, here's how to remove the extension from a filename:

```{r}
# Remove extension from filename
filename <- "My_File.txt"
idx <- unlist(gregexpr(pattern='\\.', filename))
filename_root <- substr(filename, start = 1, stop = idx[1] - 1)
print(filename_root)
```

Overwrite Characters
--------------------
Use the `str_sub()` command from the `stringr` library:

```{r}
library(stringr)

st <- "Hello World"
str_sub(st, 6, 6) <- "_"
str_sub(st, 8, 7) <- "O"
print(st)
```

Replace Characters
------------------
Use the `gsub()` command:

```{r}
st <- "Hello World"
st <- gsub("ello", "i", st)
print(st)
```

Trim White Space
----------------
*White space* is what is created by the space bar or the tab key. You can **trim** a string by removing any white space at its start and end using the `trimws()` function:

```{r}
str <- "  Hello World  "
print(trimws(str))
```

Change the Case
===============
A string can be UPPERCASE, lowercase, Sentence case or Title Case:

```{r}
st <- "The quick, brown Fox jumped over the lazy Dog"
print(str_to_upper(st))
print(str_to_lower(st))
print(str_to_sentence(st))
print(str_to_title(st))
```
These functions come from the `stringr` library.

Formatted Output
================
If you want to print a variable as part of a string using a specific format, use the `sprintf()` function (**s**tring **print** **f**ormatted) along with the "%" character to indicate where you want the variable to be inserted. You also need to have a letter after the "%" to indicate what type of format to use:

Number formats
--------------
When working with numbers, use the "f" for 'floating-point' in conjunction with "%":

```{r}
x <- 1
sprintf("x = %f", x)
```

Use a decimal point and a number to set the number of decimal places you see:

```{r}
x <- 1
sprintf("x = %.2f", x)
```

Use a number before the decimal point to set the 'width' of the number, ie how many spaces it takes up (including the decimal point). If the width is set to be wider than the number is long, the extra space with be filled up with blanks:

```{r}
x <- 1
sprintf("x = %8.2f", x)
```

String formats
--------------
Instead of the "f", use an "s" together with the "%".

You can print to the console immediately:

```{r}
st <- "Hello world"
sprintf("Output text is: %s", st)
```

...or you can write the value to a variable and print it to the console at a later stage:

```{r}
st <- "Hello world"
st <- sprintf("Output text is: %s", st)
print(st[1])
```

Using formatted output with a dataframe
---------------------------------------
If you extract a single string from any vector or dataframe, it will be treated in the same way by the `sprintf()` function:

```{r}
df <- data.frame(
    name <- c("Alfa", "Bravo", "Charlie")
)
sprintf("Output text is: %s", df$name[1])
```

Unicode and Emojii
==================
Unicode characters can be called by using `\U` and referencing their unicode number. Some can then simply be printed:

```{r}
st1 <- "\U03BC"  # mu
st2 <- "\U03B5"  # epsilon
st3 <- "\U03C9"  # omega
print(c(st1, st2, st3))
```

Others (eg emojis) cannot be printed this easily and need the special `utf8_print()` function from the `utf8` library:

```{r}
st1 <- "\U2705"  # white heavy check mark
st2 <- "\U274c"  # cross mark
st3 <- "\U0001f609"  # winking face
utf8::utf8_print(c(st1, st2, st3))
```

[⇦ Back](../../R.html)

</font>
