---
title: '<font size="5">Graphs in R using ggplot2:</font><br>Box Plots (aka Box-and-Whisker Plots) with Multiple Groups'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../../google_analytics.html
---

<font size="3">

[⇦ Back](../../../R.html)

For this page we will use data from the finals of various Olympic events, available on Wikipedia.

This page functions as a follow-on from [Boxplots with One Group of Data](boxplots_single_group.html).

Too Simple
==========
If we try to plot multiple datasets (eg the results from more than one Olympic Games) with two groups within each dataset (men and women), things start to become confusing:

```{r}
library(ggplot2)

long_jump <- data.frame(
    olympics = c(
        rep("London 2012 (Men)", 8), rep("Rio 2016 (Men)", 8),
        rep("London 2012 (Women)", 8), rep("Rio 2016 (Women)", 8)
    ),
    distance = c(
        8.31, 8.16, 8.12, 8.11, 8.10, 8.07, 8.01, 7.93,
        8.38, 8.37, 8.29, 8.25, 8.17, 8.10, 8.06, 8.05,
        7.12, 7.07, 6.89, 6.88, 6.77, 6.76, 6.72, 6.67,
        7.17, 7.15, 7.08, 6.95, 6.81, 6.79, 6.74, 6.69
    )
)

p <- ggplot(long_jump, aes(olympics, distance))
p <- p + geom_boxplot()
print(p)
```

In the above figure we just have four distinct box plots without the groups having been separated out.

A Better Approach
=================
It's a good idea to use **colour**, a **legend** and the **ordering** of the box plots to show the different data sets. This will help to make things as clear as possible:

```{r}
long_jump <- data.frame(
    olympics = c(
        rep("London 2012", 8), rep("Rio 2016", 8),
        rep("London 2012", 8), rep("Rio 2016", 8)
    ),
    men_women = c(rep("Men", 16), rep("Women", 16)),
    distance = c(
        8.31, 8.16, 8.12, 8.11, 8.10, 8.07, 8.01, 7.93,
        8.38, 8.37, 8.29, 8.25, 8.17, 8.10, 8.06, 8.05,
        7.12, 7.07, 6.89, 6.88, 6.77, 6.76, 6.72, 6.67,
        7.17, 7.15, 7.08, 6.95, 6.81, 6.79, 6.74, 6.69
    )
)

p <- ggplot(long_jump, aes(olympics, distance))
p <- p + geom_boxplot(aes(fill = men_women))
print(p)
```

Alternate Grouping
------------------
An improvement to the above graph would be to swap the grouping: plot by gender first and then by Olympics rather than Olympics first and gender second:

```{r}
long_jump <- data.frame(
    olympics = c(
        rep("Athen 2004", 8), rep("Beijing 2008", 8), rep("London 2012", 8), rep("Rio 2016", 8),
        rep("Athen 2004", 8), rep("Beijing 2008", 8), rep("London 2012", 8), rep("Rio 2016", 8)
    ),
    men_women = c(rep("Men", 32), rep("Women", 32)),
    distance = c(
        8.59, 8.47, 8.32, 8.31, 8.25, 8.24, 8.23, 8.21,
        8.34, 8.24, 8.20, 8.19, 8.19, 8.16, 8.07, 8.00,
        8.31, 8.16, 8.12, 8.11, 8.10, 8.07, 8.01, 7.93,
        8.38, 8.37, 8.29, 8.25, 8.17, 8.10, 8.06, 8.05,
        7.07, 7.05, 7.05, 6.96, 6.85, 6.83, 6.80, 6.73,
        7.04, 7.03, 6.91, 6.79, 6.76, 6.70, 6.64, 6.58,
        7.12, 7.07, 6.89, 6.88, 6.77, 6.76, 6.72, 6.67,
        7.17, 7.15, 7.08, 6.95, 6.81, 6.79, 6.74, 6.69
    )
)

p <- ggplot(long_jump, aes(men_women, distance))
p <- p + geom_boxplot(aes(fill = olympics))
print(p)
```

Adding Labels and Customising Colour
====================================
Improve the graph further by adding a title and changing the labels and colours:

- `ggtitle()` controls the graph title
- `ylab()` sets the y-axis label
- `xlab()` sets the x-axis label
- `labs()` can be used to change the title of the legend
- `scale_fill_manual()` will manually change the fill colours in both the boxes themselves and also in the corresponding icons in the legend

```{r}
p <- ggplot(long_jump, aes(men_women, distance))
p <- p + geom_boxplot(aes(fill = olympics))
p <- p + ggtitle("Long Jump Finals at the Last Four Olympic Games")
p <- p + ylab("Distance [m]")
p <- p + xlab("")
p <- p + labs(fill = "Olympics")
p <- p + scale_fill_manual(values = c("#0074c5", "#d71921", "#e400a3", "#00a650"))
print(p)
```

Annotate the Sample Size and Mean Values
========================================
It can be very useful to include summary statistics right on the plot as opposed to hidden away in a paragraph of text. This requires custom functions which then get called from within the `stat_summary()` function:

```{r}
# Custom functions
sample_size <- function(x) {
    # Sample size labels
    return(
        data.frame(y = median(x) * 1.06, label = paste0("n=", length(x)))
    )
}
sample_mean <- function(x) {
    # Sample mean labels
    return(
        data.frame(
            y = median(x) * 1.05,
            label = paste0("\U00B5=", round(mean(x), 2), "m")
        )
    )
}


p <- ggplot(long_jump, aes(men_women, distance, fill = olympics))
p <- p + geom_boxplot()
p <- p + ggtitle("Long Jump Finals at the Last Four Olympic Games")
p <- p + ylab("Distance [m]")
p <- p + xlab("")
p <- p + labs(fill = "Olympics")
p <- p + scale_fill_manual(values = c("#0074c5", "#d71921", "#e400a3", "#00a650"))
p <- p + stat_summary(
    fun.data = sample_size, geom = "text", fun = median,
    position = position_dodge(width = 0.75), size = 3
)
p <- p + stat_summary(
    fun.data = sample_mean, geom = "text", fun = median,
    position = position_dodge(width = 0.75), size = 3
)
print(p)
```

Overlaying a Scatter Plot
=========================
Instead of representing the multiple groups via multiple box plots we could use one box plot and overlay the individual points as a scatter plot. Adding colour-coding will then ensure that the sub-groups are still distinguishable:

```{r}
p <- ggplot(long_jump, aes(x = men_women, y = distance))
p <- p + geom_boxplot()
p <- p + geom_jitter(aes(color = factor(olympics)), width = 0.2)
p <- p + ggtitle("Long Jump Finals at the Last Four Olympic Games")
p <- p + ylab("Distance [m]")
p <- p + xlab("")
p <- p + labs(color = "Olympics")
p <- p + scale_color_manual(values = c("#0074c5", "#d71921", "#e400a3", "#00a650"))
print(p)
```

There are a couple of differences between this code and previous examples:

- In the `aes()` function there is no `fill` argument being used. This stops the data from being split into one box plot per Olympics.
- The `geom_jitter()` function is included. This spreads the data points in the scatter plots out to make them more visible, and the `color` argument causes them to be coloured by Olympics.
- In the `labs()` function, the label for the `color` is being set, not the `fill` as has been done previously. This is because we are using the *colour of the points* to distinguish the groups, not the *fill colour of the boxes.*
- For the same reason as above, the `scale_color_manual()` function is needed to control the colour of the points, as opposed to `scale_fill_manual()`

Save Plot
=========
Finally, save your plot to your computer as an image using one of the following (depending on what format you want the image to be in):

- `png("File Name.png")`
- `pdf("File Name.pdf")`
- `ggsave("File Name.png")`

If you use one of the first two, it must come __*before*__ you start plotting the graph (ie before you call `ggplot()`). If you use the last one (`ggsave()`) it must come __*after*__ you've plotted the graph.

[⇦ Back](../../../R.html)

</font>
