---
title: '<font size="5">Graphs in Base R:</font><br>Bar Plots (Single Factor)'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../../google_analytics.html
---

<font size="3">

[⇦ Back](../../../R.html)

Plotting Values
===============
To create a bar plot we can use the `barplot()` function in one of two ways:

1. `barplot(height)` where `height` is a **list of numbers** (eg a vector or the column of a data frame). This will create a plot where the height of each bar corresponds to its respective number in the list.
2. `barplot(formula, data)` where `formula` is a **statement** of the form `y ~ x` with y being numerical data and x being categorical data, both of which come from a data set `data`

Here is an example of each:

Numerical Data
--------------
For this example, we're going to use the built-in dataset "trees" which contains the diameter, height and volume of 31 black cherry trees (only the first 6 of which are shown here):

```{r}
print(head(trees))
```

We can use `barplot(height)` to create a bar plot where, appropriately, the height of each bar corresponds to the height of each tree:

```{r}
height <- trees$Height
barplot(height)
```

Numerical and Categorical Data
------------------------------
This example uses the built-in dataset "swiss" which has economic data from different Swiss provinces in 1888. We're going to use the 'Agriculture' column (the % of the male population working in agriculture) from six of those provinces:

```{r}
swiss$Province <- rownames(swiss)
rownames(swiss) <- 1:nrow(swiss)
print(head(swiss[c("Province", "Agriculture")]))
```

We have numerical data (`Agriculture`) and categorical data (`Province`) so we can use the second format for the `barplot()` function, namely `barplot(formula, data)`:

```{r, fig.width = 10}
barplot(Agriculture ~ Province, data = head(swiss))
```

Plotting the Frequencies of Values (Histograms)
==================================
A bar plot where the heights are the *frequencies* at which values appear in the dataset (as opposed to the values themselves) are called **histograms**. These can be plotted using the `hist()` function. Here's an example using the same `Height` data from the "trees" dataset that was used before:

```{r}
hist(trees$Height)
```

This can be changed to a *probability density* plot (where the total area of the histogram's bars is 1) by specifying `freq = FALSE`:

```{r}
hist(trees$Height, freq = FALSE)
```

Plotting the Mean Values of Groups of Data
==========================================
If we use the "chickwts" dataset we get the weights of 71 chicks, measured six weeks after hatching, that were each fed one of six different diets (first 15 data points shown):

```{r}
print(head(chickwts, 15))
```

Even though we have both categorical and numerical data we can't immediately plot it with `barplot(weight ~ feed, data = chickwts)` because *there are multiple data points in each group*. R would try to plot one bar for each row of the data frame and it would fail because multiple data points would have the same x-value ("horsebean", "linseed", etc). We first need to **aggregate** it to get, for example, the mean value of each group:

```{r}
data <- aggregate(weight ~ feed, data = chickwts, mean)
print(data)
```

Now we can plot it:

```{r}
barplot(weight ~ feed, data = data)
```

Formatting
==========
Let’s make the plot look a little better:

Titles and Labels
-----------------
- Change the bar labels with `names.arg`
- Set the x- and y-axis labels with `xlab` and `ylab`
- Set the graph title with `main`

```{r}
barplot(
    # Plot data
    weight ~ feed, data = data,
    # Bar labels
    names.arg = c("Casein", "Horsebean", "Linseed", "Meatmeal", "Soybean", "Sunflower"),
    # Axis labels
    ylab = "Weight [g]", xlab = "Feed Type",
    # Graph title
    main = "Chicken Weights By Feed Type"
)
```

Remember that you can include Unicode in your axis titles using the `\U` Unicode indicator (eg `"Pi: \U03C0"` renders as "Pi: π").

Distinguish Groups
------------------
Now that we have our labelling sorted out for the graph as a whole let's improve how we label the different groups:

### Use Colour
Colour can be added in `barplot()` with the `col` keyword argument (see [here](symbols_linetypes_colours.html#colours) for options):

```{r}
# Custom Colours
pink <- "#FB4188"
green <- "#87C94A"
blue <- "#39C2F3"
yellow <- "#FADB39"
lgrey <- "#798287"
dgrey <- "#43454C"
colours <- c(pink, green, blue, yellow, lgrey, dgrey)

# Create bar plot
barplot(
    # Plot data
    weight ~ feed, data = data,
    # Bar labels
    names.arg = c("Casein", "Horsebean", "Linseed", "Meatmeal", "Soybean", "Sunflower"),
    # Axis labels
    ylab = "Weight [g]", xlab = "Feed Type",
    # Graph title
    main = "Chicken Weights By Feed Type",
    # Add colour
    col = colours
)
```

### Use Colour *and* a Legend
- A legend can be added with the `legend()` function, specifying:
    - The location of the box as a positional argument
    - Its title with `title`
    - The text labels to display using `legend`
    - The type of markers to display in the legend with `pch` (the options for this are [here](symbols_linetypes_colours.html#symbols))
    - The colours to attach to the labels using `col`
    - We want to place the legend outside of the plot area, so we need to create an 'inset' for it to sit in. This is done using the `inset` keyword argument.
- In order for the legend to be placed outside of the plot area, we need to tweak the **par**ameters of the graph with the `par()` function:
    - Create extra space on the right for the legend to sit in by changing the **mar**gins of the graph with the `mar` argument. The default for the margins is `c(bottom, left, top, right) = c(5, 4, 4, 2) + 0.1` so we are adding 6 units of space between the right edge of the graph and the side of the figure (and we are also removing 4 units of space from beneath the plot).
    - `xpd = TRUE` allows elements (eg our legend) to be drawn outside of the plot area
- Lastly, we want to turn off the x-axis labels (as they have been moved to the legend) with `xaxt = 'n'` and `xlab = ""`

```{r}
# Custom Colours
pink <- "#FB4188"
green <- "#87C94A"
blue <- "#39C2F3"
yellow <- "#FADB39"
lgrey <- "#798287"
dgrey <- "#43454C"
colours <- c(pink, green, blue, yellow, lgrey, dgrey)
# Custom labels
labels <- c("Casein", "Horsebean", "Linseed", "Meatmeal", "Soybean", "Sunflower")

# Add extra space to the right of the plot and enable drawing outside of the plot area
par(mar = c(1.1, 4.1, 4.1, 8.1), xpd = TRUE)
# Create bar plot
barplot(
    # Plot data
    weight ~ feed, data = data,
    # Bar labels
    xaxt = 'n',
    # Axis labels
    ylab = "Weight [g]", xlab = "",
    # Graph title
    main = "Chicken Weights By Feed Type",
    # Add colour
    col = colours
)
# Add a legend
legend("right", title = "Feed Type", legend = labels, pch = 16, col = colours, inset=c(-0.3,0))
```

Show Individual Points
----------------------
This is a little tricky. In the first instance, see what happens if you simply add a scatter plot on top of the bar plot with `points()`:

```{r}
# Create bar plot
barplot(
    # Plot data
    weight ~ feed, data = data,
    # Bar labels
    names.arg = c("Casein", "Horsebean", "Linseed", "Meatmeal", "Soybean", "Sunflower"),
    # Axis labels
    ylab = "Weight [g]", xlab = "Feed Type",
    # Graph title
    main = "Chicken Weights By Feed Type"
)
# Add a scatter plot
points(chickwts$feed, chickwts$weight)
```

The data is correct but it's offset from the bars! The reason this happens is because R doesn't plot the bars at exactly x = 1, x = 2, etc. This can be seen by looking up the exact x-values R uses when creating the bars: assign the barplot to a variable and print its value:

```{r, eval = FALSE}
bp <- barplot(weight ~ feed, data = data)
print(bp)
```

```
##      [,1]
## [1,]  0.7
## [2,]  1.9
## [3,]  3.1
## [4,]  4.3
## [5,]  5.5
## [6,]  6.7
```

This has produced a matrix with six values which correspond to the x-values of each of the six bars. As you can see, they are not 1, 2, 3, etc! One way to solve this problem is to replace the x-data we want to use in our scatter plot (which, at the moment is the categorical `feed` data: casein, horsebean, linseed, etc) with these x-values. The first step to doing this is to convert the `chickwts$feed` column to numerical data:

```{r}
# Convert categorical data to numerical
chickwts$feed <- as.numeric(chickwts$feed)
print(chickwts$feed)
```

These numbers correspond to which of the six bars the data in that row will be plotted on. We can use these to look-up the actual x-value of these bars from the matrix we got earlier:

```{r, eval = FALSE}
# Replace ordinal data with the x-values of the bars
for (i in 1:nrow(chickwts)) {
    chickwts$feed[i] <- bp[chickwts$feed[i]]
}
print(chickwts$feed)
```

```
##  [1] 1.9 1.9 1.9 1.9 1.9 1.9 1.9 1.9 1.9 1.9 3.1 3.1 3.1 3.1 3.1 3.1 3.1 3.1 3.1
## [20] 3.1 3.1 3.1 5.5 5.5 5.5 5.5 5.5 5.5 5.5 5.5 5.5 5.5 5.5 5.5 5.5 5.5 6.7 6.7
## [39] 6.7 6.7 6.7 6.7 6.7 6.7 6.7 6.7 6.7 6.7 4.3 4.3 4.3 4.3 4.3 4.3 4.3 4.3 4.3
## [58] 4.3 4.3 0.7 0.7 0.7 0.7 0.7 0.7 0.7 0.7 0.7 0.7 0.7 0.7
```

Now we can plot this as a normal x-y scatter plot on top of the bar plot using `points()`. It's a good idea to increase the y-axis limits (using `ylim()`) when doing this to ensure that all of the points can fit:

```{r}
# Create bar plot
bp <- barplot(
    # Plot data
    weight ~ feed, data = data,
    # Bar labels
    names.arg = c("Casein", "Horsebean", "Linseed", "Meatmeal", "Soybean", "Sunflower"),
    # Axis labels
    ylab = "Weight [g]", xlab = "Feed Type",
    # Axis limits
    ylim = c(0, max(chickwts$weight) * 1.1),
    # Graph title
    main = "Chicken Weights By Feed Type"
)
# Convert categorical data to numerical
chickwts$feed <- as.numeric(chickwts$feed)
# Replace ordinal data with the x-values of the bars
for (i in 1:nrow(chickwts)) {
    chickwts$feed[i] <- bp[chickwts$feed[i]]
}
# Add a scatter plot
points(chickwts$feed, chickwts$weight)
```

Save Plot
=========
Finally, use `png("Name of Plot.png")` to save the plot as a PNG file to your computer.

[⇦ Back](../../../R.html)

</font>
