---
title: '<font size="5">Graphs in Base R:</font><br>Receiver Operating Characteristic Curves'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../../google_analytics.html
---

<font size="3">

[⇦ Back](../../../R.html)

This page just talks about how to _plot_ receiver operating characteristic (ROC) curves. For more on what they are and how to use them, see [this tutorial](../../statistics/tutorial.html). Briefly, a ROC curve illustrates how the diagnostic accuracy (ie the sensitivity and specificity) of a binary classification test (eg predicting whether something did or didn't happen) changes as its threshold changes.

Example Data
============
A good data set to use is the training data subset from the `titanic` package. This contains information about the passengers on board the Titanic, how much they paid for their fare and whether or not they survived. Install this package in R with:

```{r, eval=FALSE}
# Install the package
install.packages("titanic", repos = "http://cran.us.r-project.org")
```

Then load and take a look at the data:

```{r}
# Import the package
library(titanic)
# Change the width of the console
options("width" = 200)
# Take a look at the first 6 rows of the data
head(titanic_train[c("Survived", "Pclass", "Name", "Sex", "Age", "Fare", "Cabin", "Embarked")])
```

Calculating Thresholds
======================
Fortunately for us, there is a function that can do most of the work when it comes to ROC analysis: the unsurprisingly named `roc()` function. It comes with the `pROC` library which you can install via the following:

```{r, eval=FALSE}
# Install the package
install.packages("pROC", repos = "http://cran.us.r-project.org")
```

Then load and use:

```{r message = FALSE}
library(pROC)

r <- roc(Survived ~ Fare, data = titanic_train)
```

Have a look at some of the thresholds (ie values for fare that will be taken as cut-offs) that this function produced:

```{r}
head(r$thresholds)
```

Plot
====

```{r fig.height = 5, fig.width = 5}
plot(r, type = "S")
```

Get the area under the curve and the 95% confidence interval of this area (which should be rounded off to 2 decimal places):

```{r}
auc <- auc(r)
ci <- ci.auc(r)
ci_l <- round(ci[1], 2)
ci_u <- round(ci[3], 2)
```

Add this information to the graph:

```{r fig.height = 5, fig.width = 5}
plot(r, type = "S")
legend_text <- paste0(
    "AUC = ", round(auc, 2), " (95% CI = ", ci_l, " - ", ci_u, ")"
)
legend("bottomright", legend = legend_text, pch = 15)
```

The area under the curve is 0.69; not particularly high! It seems that fare can't be used as an accurate predictor of passenger survival.

[⇦ Back](../../../R.html)

</font>
