---
title: '<font size="5">Graphs in R using plot():</font><br>Symbols, Line Types and Colours'
output:
    html_document:
        theme: paper
        highlight: textmate
        number_sections: true
        toc: true
        includes:
            before_body: ../../../google_analytics.html
---

<font size="3">

[⇦ Back](../../../R.html)

Symbols
=======
By default, there are 26 symbols that can be used for the data points of a scatter plot. These are set using the `pch` keyword argument and are numbered from 0 to 25, for example:  
`plot(x, y, pch = 0)`  
will plot with squares while  
`plot(x, y, pch = 1)`  
will plot with circles, and so on.

```{r out.width="100%"}
num_symbols <- 26
nrows <- 13
ncols <- 2
# Set the margins of the plotting area
par(mar = c(0, 0, 0, 0))
# Plot
x <- c()
y <- c()
for (i in 0:25) {
    x <- c(x, ceiling((i + 1) / nrows))
    y <- c(y, nrows - (i %% nrows))
}
plot(x, y, pch = 0:25, axes = FALSE, xlim = c(0.75, ncols + 0.75))
name <- c(
    "square", "circle", "triangle point up", "plus", "cross", "diamond",
    "triangle point down", "square cross", "star", "diamond plus",
    "circle plus", "triangles up and down", "square plus", "circle cross",
    "square and triangle down", "filled square", "filled circle",
    "filled triangle point-up", "filled diamond", "solid circle",
    "bullet (smaller circle)", "filled circle blue", "filled square blue",
    "filled diamond blue", "filled triangle point-up",
    "filled triangle point dwn"
)
text(x + 0.1, y, name, adj = 0)
text(x - 0.1, y, 0:25)
```

Line Types
==========
The style of line that is plotted can be controlled with the `lty` keyword argument (**l**ine **ty**pe), for example:  
`plot(x, y, lty = "dashed", type = "l")`  
will plot a dashed line. Note that the `type` keyword argument needs to be set to `l` for a line to be plotted (only the points will be plotted by default - ie a scatter plot instead of a line plot). The line types that can be used are as follows:

```{r, message=FALSE}
library(ggpubr)
show_line_types()
```

Colours
=======
There are 657 colours that can be referred to by name when choosing a colour for a plot. These are set using the `col` keyword argument, for example:  
`plot(x, y, col = "red")`  
will plot in red. The complete list is below:

```{r out.width="100%", fig.height=50, fig.width=15}
num_colours <- 657
nrows <- 110
ncols <- 6
# Set the margins of the plotting area
par(mar = c(0, 0, 0, 0))
# Plot
plot(
    seq(1.5, 3.5, by = 2), seq(1.5, 3.5, by = 2),
    xlim = c(0, ncols * 2), ylim = c(0, nrows),
    xlab = NULL, ylab = NULL, xaxt = "n", yaxt = "n", axes = FALSE, ann = FALSE
)
xleft <- rep(seq(1, ncols * 2 - 1, by = 2), nrows)[1:num_colours]
ybottom <- c()
for (i in nrows:1 - 1) {
    ybottom <- c(ybottom, rep(i, ncols))
}
ybottom <- ybottom[1:num_colours]
xright <- rep(seq(1.5, ncols * 2 - 0.5, by = 2), nrows)[1:num_colours]
ytop <- c()
for (i in nrows:1) {
    ytop <- c(ytop, rep(i, ncols))
}
ytop <- ytop[1:num_colours]
rect(xleft, ybottom, xright, ytop, col = colors()[1:num_colours])
xtext <- rep(seq(0.25, ncols * 2 - 1.75, by = 2), nrows)[1:num_colours]
ytext <- c()
for (i in nrows:1 - 0.5) {
    ytext <- c(ytext, rep(i, ncols))
}
ytext <- ytext[1:num_colours]
text(xtext, ytext, colors()[1:num_colours], cex = 1.2, font = 2)
```

Transparency
============
The easiest way to edit the transparency of an element is to use the `rgb()` function when setting its colour. Instead of `col = "colour_name"` as is being demonstrated above, this would involve using `col = rgb()` and specifying the amounts of red, green and blue (on scales of 0 to 1) in the colour you want. The advantage of this method is that `rgb()` also has an `alpha` keyword argument which, when given a number between 0 and 1, sets the transparency of the colour:

```{r}
height <- c(1, 1, 1, 1, 1, 1)
alpha <- c(0.0, 0.2, 0.4, 0.6, 0.8, 1.0)
barplot(height, col = rgb(red = 1, green = 0, blue = 0, alpha = alpha), yaxt = "n")
```

Marker Size
===========
This is a **par**ameter of the plot and so it needs to be set with the `par()` function. Specifically, use the `cex` keyword argument as you are changing the **c**haracter **ex**pansion relative to the default (which is 1):

```{r}
par(cex = 1)
x <- c(1, 2, 3, 4, 5)
y <- c(1, 1, 1, 1, 1)
plot(x, y, ylim = c(1, 3))

par(cex = 2)
x <- c(1, 2, 3, 4, 5)
y <- c(2, 2, 2, 2, 2)
points(x, y)

par(cex = 3)
x <- c(1, 2, 3, 4, 5)
y <- c(3, 3, 3, 3, 3)
points(x, y)
```

[⇦ Back](../../../R.html)

</font>
